---
title: "Chapter 3"
output: html_notebook
---


Loading libraries, getting play by play data and filtering for run plays. Setting all null value runs to 0.
```{r}
library(tidyverse)
library(nflfastR)

playByPlay = load_pbp(2016:2023)

runningPlays =
        playByPlay |>
                filter(play_type == "run" & !is.na(rusher_id))|>
                mutate(rushing_yards = ifelse(is.na(rushing_yards), 0, rushing_yards))
```



Constructing some plots
```{r}
ggplot(runningPlays, aes(x=ydstogo, y=rushing_yards)) +
        geom_point() +
        theme_bw() +
        stat_smooth(method="lm")

```
Yards to go plotted against rushing yards
We can see that rushing yards increase as yards to go increases, albeit very slightly.


```{r}
runAverage =
        runningPlays |>
                group_by(ydstogo) |>
                summarize(ypc = mean(rushing_yards))

ggplot(runAverage, aes(x = ydstogo, y = ypc)) +
        geom_point() +
        theme_bw() +
        stat_smooth(method="lm")
```
Average yards per car by yards to go. There is a positive linear relationship between yards per carry and yards to go.



```{r}
yardsToGo = lm(rushing_yards ~ 1 + ydstogo, data= runningPlays)
summary(yardsToGo)
```
```{r}
runningPlays <-
        runningPlays |>
                mutate(ryoe = resid(yardsToGo))

ryoe <-
        runningPlays |>
                group_by(season, rusher_id, rusher) |>
                summarize(
                  n = n(),
                  ryoe_total = sum(ryoe),
                  ryoe_per = mean(ryoe),
                  yards_per_carry = mean(rushing_yards)
                ) |>
                arrange(-ryoe_total) |>
                filter(n > 50)

print(ryoe)
```




```{r}
ryoe |>
        arrange(-ryoe_per)

```



```{r}
ryoe_now =
        ryoe |>
                select(-n, -ryoe_total)

ryoe_last =
        ryoe |>
                select(-n, -ryoe_total) |>
                mutate(season = season + 1) |>
                rename(ryoe_per_last = ryoe_per, yards_per_carry_last = yards_per_carry)

ryoe_lag =
        ryoe_now |>
                inner_join(ryoe_last, by = c("rusher_id", "rusher", "season")) |>
                ungroup()

ryoe_lag |>
        select(yards_per_carry, yards_per_carry_last) |>
        cor(use = "complete.obs")

ryoe_lag |>
        select(ryoe_per, ryoe_per_last) |>
        cor(use = "complete.obs")
```
```{r}



```

