---
title: "R Notebook"
output: html_notebook
---

The [R plugin](https://www.jetbrains.com/help/pycharm/r-plugin-support.html) for IntelliJ-based IDEs provides
handy capabilities to work with the [R Markdown](https://www.jetbrains.com/help/pycharm/r-markdown.html) files.
To [add](https://www.jetbrains.com/help/pycharm/r-markdown.html#add-code-chunk) a new R chunk,
position the caret at any line or the code chunk, then click "+".

The code chunk appears:
```{r}
library(nflfastR)
library(tidyverse)
```

Load in the data and filter for passing plays
```{r}
playByPlay =
        load_pbp(2016:2023)

passingPlay =
        playByPlay |>
                filter(!is.na(passer_id))
```


Next we will repalce the NUll or NA values with 0 for pass_touchdown, then we will aggregate by season, week, passer_id, and passer to calculate the number of passes per week and the number of tocuhdown passes per week. We will exclude players will less than 10 plays as a passer for each week Last, we'll save the total_line
```{r}
passing_tds =
        passingPlays |>
                mutate(pass_touchdown = ifelse(is.na(pass_touchdown), 0, pass_touchdown)) |>
                group_by(season, week, passer_id, passer) |>
                summarize(
                  n=n(),
                  pass_td_y = sum(pass_touchdown),
                  total_line = mean(total_line)
                ) |>
                filter(n >=10)

passing_tds |>
        group_by(pass_td_y) |>
        summarize(n = n())
```




```{r}
passing_tds |>
        ungroup() |>
        select(-passer, -passer_id) |>
        summary()
```


We want to see if poisson is a reasonable distribution, so we look at a histogram of the frquencies and compare it with a Poisson distribution

```{r}
pass_td_mean =
        passing_tds |>
                pull(pass_td_y) |>
                mean()

plot_pos =
        tibble(x = seq(0, 7)) |>
                mutate(expected = dpois(
                  x = x, lambda = pass_td_mean
                ))

ggplot() +
        geom_histogram(
          data=passing_tds,
          aes(
            x=pass_td_y,
            y=after_stat(count / sum(count))
          ),
          bindwidth = 0.5
        ) +
        geom_line(
          data=plot_pos, aes(x =x ,y= expected), color = "red", linewidth=1
        ) +
        theme_bw() +
        xlab("Touchdown passes per player per game for 2016 to 2023") +
        ylab("Probability")









```






```{r}
pass_td_geq10 =
        passing_tds |>
                filter(n >= 10)
x = tibble()

for (season_idx in seq(2017, 2023)) {
  for(week_idx in seq(1,22)) {
    week_calc =
      pass_td_geq10 |>
        filter((season == (season_idx -1)) |
                 (season == season_idx & week < week_idx)) |>
        group_by(passer_id, passer) |>
        summarize(
          n_games = n(),
          pass_td_rate = mean(pass_td_y),
          .groups = "keep"
        ) |>
        mutate(season = season_idx, week = week_idx)
    x = bind_rows(x, week_calc)
  }
}





```






```{r}
x |>
        filter(passer == "P.Mahomes") |>
        tail()
```
The books have set a decent number on Mahomes passing td totals


We will create a response varaible which is the dataframe we created along with the added game total
```{r}
pass_td_geq10 =
        pass_td_geq10 |>
                inner_join(x, by = c("season", "week", "passer_id", "passer"))

pass_td_geq10
```




We plot passing touchdowns in each game for each passer by using a line


```{r}
weekly_passing_id_plot =
        pass_td_geq10 |>
                ggplot(aes(x = week, y = pass_td_y, group=passer_id)) +
                geom_line(alpha = 0.25) +
                facet_wrap(vars(season), nrow = 3) +
                theme_bw() +
                theme(strip.background = element_blank()) +
                ylab("Total passing touchdowns") +
                xlab("Week of season")

weekly_passing_id_plot
```

This shows the variability in passing touchdowns per game. The values seem to be constant and there's no obvious trend. We now add a Poisson regression lin


```{r}
weekly_passing_id_plot +
        geom_smooth(method='glm', method.args = list("family" = "poisson"), se = FALSE, linewidth = 0.5, color ='blue', aplha=0.25)
```




We now fit the model, save the outputs and look at the summary

```{r}
pass_fit =
        glm(pass_td_y ~ pass_td_rate.x + total_line, data = pass_td_geq10, family = "poisson")

pass_td_geq10 =
  pass_td_geq10 |>
    ungroup() |>
    mutate(exp_pass_td = predict(pass_fit, type="response"))

summary(pass_fit)|>
  print()


```



```{r}
library(broom)
tidy(pass_fit, exponentiate = TRUE, conf.int = TRUE)
```



```{r}
pass_td_geq10 |>
        filter(passer == "P.Mahomes", season == 2022, week ==22) |>
        select(-pass_td_y, -passer_id, -week, -season, -n_games.x)
```


WE use the dpois() function to give the probability of zero, one or two touchdown passes

```{r}
pass_td_geq10 =
        pass_td_geq10 |>
                mutate(
                  p_0_td = dpois(x = 0, lambda = exp_pass_td),
                  p_1_td = dpois(x = 1, lambda = exp_pass_td),
                  p_2_td = dpois(x = 2, lambda = exp_pass_td),
                  p_g2_td = ppois(q = 2, lambda = exp_pass_td, lower.tail = FALSE)
                )

pass_td_geq10 |>
        filter(passer == "P.Mahomes", season == 2022, week == 22) |>
        select(-pass_td_y, -n_games.x, -n, -passer_id, -week, -season)
```


```{r}
x = rpois(n = 10, lambda = 1)
print(x)
print(mean(x))
```


```{r}
library(broom)
glm_out = glm(x ~1, family = "poisson")

print(tidy(glm_out))
```

```{r}
print(tidy(glm_out, exponentiate = TRUE))
```


```{r}
bal_td =
        playByPlay |>
                filter(posteam == "BAL" & season == 2022) |>
                group_by(game_id, week) |>
                summarize(
                  td_per_game = sum(touchdown, na.rm = TRUE),
                  .groups = "drop"
                ) |>
                mutate(week = week -1)

ggplot(bal_td, aes(x = week, y = td_per_game)) +
        geom_point() +
        theme_bw() +
        stat_smooth(
          method = "glm", formula = "y~x", method.args = list(family = "poisson")
        ) +
        xlab("Week") +
        ylab("Touchdowns per game") +
        scale_y_continuous(breaks = seq(0, 6)) +
        scale_x_continuous(breaks = seq(1, 20, by = 2))







```


```{r}
glm_bal_td =
        glm(td_per_game ~ week, data = bal_td, family = "poisson")

print(tidy(glm_bal_td))
print(tidy(glm_bal_td, exponentiate = TRUE))
```










